---
title: Economic and Life-Threatening Events as a result of Storm Activities in the
  USA
author: "Alan C Bonnici"
date: "Monday, August 18, 2014"
output:
  html_document:
    keep_md: yes
---
```{r supportFunctions, echo=FALSE, results='hide'}
cleanEVTYPE <- function(df) {
    # unfactor the EVTYPE
    df$EVTYPE <- as.character(df$EVTYPE)
    
    srch <- c("COASTAL SURGE|MIXED PRECIP|SEICHE|DOWNBURST|WETNESS|STREAM|PRECIPITATION|SHOWER|TIDE|RAIN|DROWNING|FLOOD|SURF|STORM|WAVE|WATER|CURRENTS|WATERSPOUT|MARINE|SEA|SWELL|STREAM FLD|CURRENT",
              "AVAL",
              "ICE|SNOW|BLIZZARD|FREEZ|FROST|HAIL|HYPOTHERMIA|COLD|LOW TEMPERATURE|SLEET|ICY|HYPERTHERMIA|WINTRY|WINTER",
              "DUST|WIND|TYPHOON|TORNADO|HURRICANE|FUNNEL|MICROBURST|GUSTNADO|TORNDAO|TROPICAL|TURBULENCE|LANDSPOUT",
              "FOG|DENSE SMOKE",
              "HEAT|WARM|HOT|DROUGHT",
              "FIRE",
              "LIGHTING|LIGHTNING|TSTM|LIGHTING",
              "MUD|LANDSLIDES|EROSION|ROCK SLIDE|VOLCANIC ASH|LANDSLUMP|LANDSLIDE|TSUNAMI",
              "HIGH|APACHE COUNTY|DAM BREAK|URBAN|GLAZE|HEAVY MIX|COOL AND WET|\\?")
    
    repl <- c("WATER",
              "AVALANCHE",
              "WEATHER-COLD",
              "WIND",
              "VISIBILITY",
              "WEATHER-HOT",
              "FIRE",
              "LIGHTING",
              "LAND",
              "OTHER")
    
    for (i in 1:length(srch)) {
        x <- grepl(srch[i], df$EVTYPE, ignore.case = TRUE)
        df[x, "EVTYPE"] <- repl[i]
    }
    
    df$EVTYPE <- as.factor(df$EVTYPE)
    return (df)
}
```

```{r loadLibraries, echo=FALSE, results='hide'}
# Install and load libraries used here
for (x in c("ggplot2","reshape")) {
    if (x %in% rownames(installed.packages()) == FALSE) {
        install.packages(x)
        }
    }
library(ggplot2)
library(reshape)
```

## Introduction

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

### Assignment

The purpose of this report is to answer the following questions:

1. Across the United States, which types of events are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?


### Data

``` {r init, echo=FALSE, results='hide'}
# create the directory if it does not exists. 
if (!file.exists("./data")) {
    dir.create("./data")
}

zipFileNm <- "./data/StormData.csv.bz2"
dLoadFileNm <- "./data/sourceDt.RData"

# if the zip file does not exist download the file
if (!file.exists(zipFileNm)) {
    # this file stores the date when the archive was downloaded
    
    url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
    download.file(url, destfile = zipFileNm, method="curl")
    
    # read the data directly from the archive
    data <- read.csv(bzfile(zipFileNm))
    
    # update the date the file was downloaded
    downloadDtTm <- format(Sys.time(), "%b %d %Y at %H:%M:%S")
    save(downloadDtTm, data, file = dLoadFileNm)        
} else {
    # rather than process all the data load presaved environment
    load(dLoadFileNm, envir=environment())    
}
```

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

The data used is this analysis was downloaded from https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2. It was downloaded on **`r downloadDtTm`**.

```{r DateCleanup, echo=FALSE, results='hide'}
data$BGN_DATE <- strptime(paste(as.Date(data$BGN_DATE, format="%m/%d/%Y"), data$BGN_TIME), format="%Y-%m-%d %H%M")
data$BGN_TIME <- NULL

# Convert certain columns to upper case
data$EVTYPE <- toupper(data$EVTYPE)
data$PROPDMGEXP <- toupper(data$PROPDMGEXP)
data$CROPDMGEXP <- toupper(data$CROPDMGEXP)
```

The data on which this report is based consists of **`r formatC(nrow(data), format="d", big.mark=",")`** observations. The period covered by the data is between **`r format(min(data$BGN_DATE), "%d %b %Y")`** and **`r format(max(data$BGN_DATE), "%d %b %Y")`**.

\newpage

##Across the United States, which types of events are most harmful with respect to population health?

```{r ExtractInjCas}
data2 <- data[data$INJURIES > 0 | data$FATALITIES>0,c("EVTYPE", "FATALITIES", "INJURIES")]
```

Procedure: a subset of the original data was extracted from the original source. This consisted of only those rows that reported INJURIES or FATALITIES greater than zero. The resulting subset consisted of **`r formatC(nrow(data2), format="d", big.mark=",")`** observations. 

The Event Type description (*EVTYPE*) was not coded against a fixed list. This meant that there were many items that overlapped and this would confound the results. The original list of reported Event Types in the subset was the following:

``` {r UniqueEVTYPEInjFat}
unique(data2$EVTYPE)
```

After the list was consolidated, the number of groupings are the following:

```{r cleanInjFat}
data2 <- cleanEVTYPE(data2)
unique(data2$EVTYPE)
```

```{r PlotInjFat, echo=FALSE}
data2 <- melt(data2, id=c("EVTYPE"))
names(data2) <- c("EventType","Category","Amount")
p <- ggplot(data2, aes(x=EventType, y=Amount,fill=Category))
p <- p + geom_bar(stat="identity") + theme(axis.text.x = element_text(angle = 90, hjust = 1))
p + ggtitle("Harmful effects to population health by Event Type") + xlab("Event Category") + ylab("Cost US$")
```

```{r FindMaxInjFat, echo=FALSE,results='hide'}
maxs <- aggregate(Amount ~ EventType, data = data2, FUN = max)
maxr <- order(maxs$Amount, decreasing = TRUE)[1]
```


From the chart it is clearly visble that the event that costs the most in terms of harmful events to humans are those classified under **`r maxs[maxr, "EventType"]`**.

The table of the recategorised data is shown hereunder

```{r FindSumInjFat, echo=FALSE}
sums <- aggregate(Amount ~ EventType, data = data2, FUN = sum)
sums[order(sums$Amount),]
```
\newpage

##Across the United States, which types of events have the greatest economic consequences?

```{r ExtractPropCrop, echo=FALSE}
# Adjust the multiplier for the CROP Damage so that it is a number. Clear out erroneus data
data$CROPDMGEXP <- as.character(data$CROPDMGEXP)
data$CROPDMGEXP[data$CROPDMGEXP == "" | data$CROPDMGEXP == "?"] <- 0
data$CROPDMGEXP[data$CROPDMGEXP == "M"] <- 6
data$CROPDMGEXP[data$CROPDMGEXP == "K"] <- 3
data$CROPDMGEXP[data$CROPDMGEXP == "B"] <- 9

# Adjust the multiplier for the PROPERTY Damage so that it is a number. Clear out erroneus data
data$PROPDMGEXP <- as.character(data$PROPDMGEXP)
data$PROPDMGEXP[data$PROPDMGEXP == "" | data$PROPDMGEXP == "?" | data$PROPDMGEXP == "+" | data$PROPDMGEXP == "-"] <- 0
data$PROPDMGEXP[data$PROPDMGEXP == "M"] <- 6
data$PROPDMGEXP[data$PROPDMGEXP == "K"] <- 3
data$PROPDMGEXP[data$PROPDMGEXP == "B"] <- 9
data$PROPDMGEXP[data$PROPDMGEXP == "H"] <- 2

# Compute the dollar amount for the two type of damage
data$CropDmgTot <- data$CROPDMG * 10 ^ as.numeric(data$CROPDMGEXP)
data$PropDmgTot <- data$PROPDMG * 10 ^ as.numeric(data$PROPDMGEXP)

# Build the multiplier for the damage extimate
data3 <- data[data$CropDmgTot > 0 | data$PropDmgTot > 0,c("EVTYPE", "CropDmgTot", "PropDmgTot")]
```

Procedure: a subset of the original data was extracted from the original source. This consisted of only those rows that reported INJURIES or FATALITIES greater than zero. The resulting subset consisted of **`r formatC(nrow(data3), format="d", big.mark=",")`** observations. 

The Event Type description (*EVTYPE*) was not coded against a fixed list. This meant that there were many items that overlapped and this would confound the results. The original list of reported Event Types in the subset was the following:

``` {r UniqueEVTYPEPropCrop}
unique(data3$EVTYPE)
```

After the list was consolidated, the number of groupings are the following:

```{r cleanPropCrop}
data3 <- cleanEVTYPE(data3)
unique(data3$EVTYPE)
```

```{r PlotPropCrop, echo=FALSE}
data3 <- melt(data3, id=c("EVTYPE"))
names(data3) <- c("EventType","Category","Amount")
p <- ggplot(data3, aes(x=EventType, y=Amount, fill=Category))
p <- p + geom_bar(stat="identity") + theme(axis.text.x = element_text(angle = 90, hjust = 1))
p + ggtitle("Economic Costs by Event Type") + xlab("Event Category") + ylab("Cost US$")
```

```{r FindMaxPropCrop, echo=FALSE,results='hide'}
maxs <- aggregate(Amount ~ EventType, data = data3, FUN = max)
maxr <- order(maxs$Amount, decreasing = TRUE)[1]
```


From the chart it is clearly visble that the event that costs the most in terms of harmful events to humans are those classified under **`r maxs[maxr, "EventType"]`**.

The table of the recategorised data is shown hereunder


```{r FindSumPropCrop, echo=FALSE}
sums <- aggregate(Amount ~ EventType, data = data3, FUN = sum)
sums[order(sums$Amount),]
```
\newpage

## Reproducable Research

In line with good data sciences practice the code and data from which this report has been generated are available for download by anyone who would like to verify the data and the algortihms used to get to the answers. 

This is available at https://github.com/chribonn/RepData_PeerAssessment2.
